Systemix Article: Combined MAG-SMAG Profiling

### 2024-05-17 R/paper.Rmd
updated to manuscript version 1.1

suppressMessages(library(ape)) #tree basics
suppressMessages(library(treeio)) #join trees
suppressMessages(library(ggtree)) #tree plotting
suppressMessages(library(ggplot2)) #plots
suppressMessages(library(ggtreeExtra)) #geom_fruit
suppressMessages(library(ggnewscale)) #scales in ggplot2
suppressMessages(library(ggstar)) #geom_star for ggplot2
suppressMessages(library(tidyverse)) #strings, dataframes processing
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(tidytree)) #trees as dataframes
suppressMessages(library(RColorBrewer)) #color palettes
suppressWarnings(library(colorspace)) #colors manipulation
suppressWarnings(library(patchwork)) #side-by-side plotting
suppressWarnings(library(openxlsx)) #read excel
suppressWarnings(library(ComplexHeatmap)) #heatmap
suppressWarnings(library(PCAtools)) #pca
suppressWarnings(library(ggbiplot)) #pca plots
suppressWarnings(library(agricolae)) #fisher's test
suppressWarnings(library(ggpubr)) #stats & plots
suppressWarnings(library(dndscv)) #dN/dS of snps
suppressWarnings(library(clusterProfiler)) #gene enrichment
```

@@

#fig sup 1
fig_s1 <- { plot_depth_top_phyla |
    (plot_depth_usgbs / plot_depth_spec + plot_layout(heights = c(0.85, 0.15))) } +
  plot_layout(ncol = 2, widths = c(0.25, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
fig_s1[[2]][[1]] <- fig_s1[[2]][[1]] + theme(axis.title.y = element_text(vjust = 0))
#fig_s1[[2]][[2]] <- fig_s1[[2]][[2]] + theme(axis.text.y = element_text(vjust = 0))
ggsave("fig_s1.pdf", fig_s1, height = 8, width = 16)
#ggsave("fig_s1.png", fig_s1, height = 8, width = 16)
```

@@

# Sort ecosystems (by median)
anova_sort_cu <- kegg_bubble_cu %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(n = median(count_norm)) %>%
  dplyr::arrange(desc(n)) %>%
  as.data.frame()
anova_sort_fe <- kegg_bubble_fe %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(n = median(count_norm)) %>%
  dplyr::arrange(desc(n)) %>%
  as.data.frame()
anova_sort_s <- kegg_bubble_s %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(n = median(count_norm)) %>%
  dplyr::arrange(desc(n)) %>%
  as.data.frame()

@@

## Figure 2

```{r fig2}
# Merge plots using patchwork
fig_2 <- { { anova_boxplot_cu / anova_boxplot_fe +
    plot_layout(ncol = 1, widths = c(0.75)) } | anova_boxplot_s / kegg_bubbles_plot } +
  plot_annotation(tag_levels = list(c("a", "b", "c", "d"))) &
  theme(plot.tag =  element_text(face = "bold"))
#adjust y-label margin for sulfur plot
fig_2[[2]][[1]] <- fig_2[[2]][[1]] + theme(axis.title.y = element_text(vjust = -127))
ggsave("fig_2.pdf", fig_2, height = 14, width = 18)
#ggsave("fig_2.png", fig_2, height = 14, width = 18) vjust=120
```

@@

```{r snps}
# Get CDS annotations (pre-processed from Bakta results)
snp_refcds <- read.csv("mags-all/5-SNPs/dndscv/buildref/cds_table.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Buildref testing for bacteria.. (1 bin)
#https://github.com/im3sanger/dndscv/issues/62
#buildref(cdsfile = "mags-all/5-SNPs/dndscv/buildref/cds_node_25854.txt",
#         genomefile = "mags-all/5-SNPs/dndscv/buildref/node_25854.fa",
#         outfile = "mags-all/5-SNPs/dndscv/buildref/testref.rda")
#load("mags-all/5-SNPs/dndscv/buildref/testref.rda")

# Buildref for all bins (very slow: ~2 hrs in building object)
#buildref(cdsfile = "mags-all/5-SNPs/dndscv/buildref/cds_table.txt",
#         genomefile = "mags-all/5-SNPs/dndscv/buildref/S15_bins.fa",
#         outfile = "mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda")

@@

# Explore output object..
#https://github.com/im3sanger/dndscv/issues/6
#https://github.com/im3sanger/dndscv/issues/64
head(snp_dndsout$sel_cv) #dN/dS ratios at gene-level
head(snp_dndsout$annotmuts)
print(snp_dndsout$globaldnds) #driver globals

# Save dN/dS per-gene matrix
#driver discovery or genes under positive selection (PS)
snp_selcv <- snp_dndsout$sel_cv
rm(snp_dndsout)

# Group global dN/dS ratios:
#https://github.com/im3sanger/dndscv/issues/63
snp_selcv %>% dplyr::filter(pmis_cv < 0.05) %>% nrow() #n=105328
snp_global_ps <- snp_selcv %>%
  dplyr::filter(wmis_cv > 1) %>%
  dplyr::summarise(genes = n())
snp_global_ps$label = "dN/dS > 1"
snp_global_ns <- snp_selcv %>%
  dplyr::filter(wmis_cv < 1 & wmis_cv != 0) %>%
  dplyr::summarise(genes = n())
snp_global_ns$label = "dN/dS < 1"
snp_global_dnds <- rbind(snp_global_ps, snp_global_ns)
snp_global_dnds <- snp_global_dnds %>%
  dplyr::mutate(perc = round(genes * 100 / nrow(snp_selcv), 2)) %>% #add %
  dplyr::mutate(perc2 = paste0(genes, " (", perc, "%)")) #label to plot
rm(snp_global_ps, snp_global_ns) #clean memory

@@

@@

## FUNCTIONS ##
# Function that returns the contig where a gene was found
fx_dndsGene2Contig <- function(g) {
  c <- snp_refcds %>%
    dplyr::filter(gene.name == g) %>%
    dplyr::select(chr) %>%
    as.character()
  return(c)
}
# Function that returns the bin given a contig
fx_dndsContig2Bin <- function(c) {
  b <- snp_bins %>%
    dplyr::filter(contig == c) %>%
    dplyr::select(bin) %>%
    as.character()
  return(b)
}
# Function that gets gene names given an id
fx_dndsGeneName <- function(g) {
  g <- snp_genids %>%
    dplyr::filter(gene.id == g) %>%
    dplyr::select(gene.name) %>%
    as.character()
  return(g)
}
# Function that returns a row with the taxonomic classification of a bin
fx_dndsTaxaBin <- function(b) {
  t <- mag_samples %>%
    dplyr::mutate(bin = paste0(mag_name, ".", mag_id)) %>%
    dplyr::filter(bin == b) %>%
    dplyr::select(phylum, genus, species, predicted_genes)
  return(t)
}
# Function that process the product tag of NS genes
fx_dndsNsPathway <- function(p) {
  if (grepl("copper", p, ignore.case = TRUE) == TRUE)
    return("Copper")
  if (grepl("iron", p, ignore.case = TRUE) == TRUE)
    return("Iron")
  if (grepl("nickel", p, ignore.case = TRUE) == TRUE)
    return("Nickel")
  if (grepl("sulfur", p, ignore.case = TRUE) == TRUE)
    return("Sulfur")
  if (grepl("nitrogen", p, ignore.case = TRUE) == TRUE)
    return("Nitrogen")
}
# Function that gets dndscv stats for NS genes
fx_dndsNsStats <- function(g) {
  s <- snp_selcv %>%
    dplyr::filter(gene_name == g) %>%
    dplyr::select(gene_name, wmis_cv, pmis_cv)
  return(s)
}
## END FUNCTIONS ##

@@

snp_taxabins <- tibble(phylum = character(), genus = character(), species = character(),
                       completeness = numeric(), contamination = numeric())
for (i in 1:nrow(snp_dnds)) {
  #print(sprintf("%i -> bin: %s", i, fx_dndsTaxaBin(snp_dnds$bin[i])))
  snp_taxabins <- rbind(snp_taxabins, fx_dndsTaxaBin(snp_dnds$bin[i]))
}
snp_dnds['phylum'] <- snp_taxabins$phylum
snp_dnds['genus'] <- snp_taxabins$genus
snp_dnds['species'] <- snp_taxabins$species
snp_dnds['predicted_genes'] <- snp_taxabins$predicted_genes
rm(snp_taxabins)

@@

# Subset PS genes to plot mutations -> synonymous/missense (non-syn.)/nonsense
#discard HPs (by name)
#genes with at least 5 SNPs
#mark genes reaching significance (q < 0.05)
#mark genes: KJCAEPPB_04437 -> nickel pathway; NEBJFEFK_01910 -> sulfur pathway
snp_psgenes_sub <- snp_psgenes %>%
  dplyr::filter(nchar(name) < 14) %>%
  dplyr::mutate(snps = n_syn + n_mis + n_non) %>%
  dplyr::filter(snps >= 4) %>%
  dplyr::mutate(signif = ifelse(pmis_cv < 0.05, "*", "")) %>%
  dplyr::mutate(path = case_when(gene_name == "KJCAEPPB_04437" ~ "Ni",
                                 gene_name == "NEBJFEFK_01910" ~ "S", TRUE ~ "")) %>%
  distinct(name, .keep_all = TRUE) %>% #remove repeated names
  pivot_longer(cols = c(n_syn, n_mis, n_non), names_to = "mut", values_to = "n_mut")

# Save PS gene names and ids to file
if (file.exists("mags-all/5-SNPs/dndscv/results/ps_genes.txt") == FALSE) {
  write.table(snp_psgenes_sub %>% dplyr::select(gene_name) %>% distinct(gene_name), "mags-all/5-SNPs/dndscv/results/ps_genes.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}
if (file.exists("mags-all/5-SNPs/dndscv/results/kofam/ps_genes_ids.txt") == FALSE) {
  write.table(snp_psgenes_sub %>% dplyr::select(name) %>% distinct(name), "mags-all/5-SNPs/dndscv/results/kofam/ps_genes_ids.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}

# Barplot with global dN/dS ratios
snp_plot_dndsglob <- ggplot(snp_global_dnds, aes(x = genes, y = reorder(label, genes), fill = label)) +
  geom_col(position = "dodge2", fill = "#4c6194") +
  geom_text(aes(label = perc2), vjust = 0.5, hjust = -0.1, size = 2) +
  scale_x_continuous(limits = c(0, 16000)) +
  labs(x = "Number of genes", y = "") +
  theme_classic() +
  theme(legend.position = "none")
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_global.pdf", snp_plot_dndsglob, height = 2, width = 5)

# Plot global driver genes by species
snp_plot_psbins <- ggplot(snp_dnds, aes(x = num, y = reorder(bin, num), fill = bin)) +
  geom_bar(stat = "identity", fill = "#E31A1C") +
  geom_text(aes(label = num), vjust = 0.5, hjust = -0.5, size = 2.5) +
  labs(x = "Number of genes under PS", y = "Bin") +
  scale_x_continuous(limits = c(0, 255)) +
  theme_classic() +
  theme(strip.background = element_blank()) +
  facet_grid(.~species)
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psbins.pdf", snp_plot_psbins, height = 8.5, width = 9.5)

@@

snp_plot_psgenes <- (snp_plot_psgenes | snp_plot_muts) +
  plot_layout(nrow = 1, ncol = 2, widths = c(0.55, 0.5))
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psgenmuts.pdf", snp_plot_psgenes, height = 11, width = 11)
rm(snp_plot_muts)

# Load negative selected (NS) genes
snp_nsgenes <- read.csv("mags-all/5-SNPs/dndscv/results/selected_cds_metals.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Infer pathway from product and add dnds attributes
snp_nspath <- tibble(gene_name = character(), pathway = character())
snp_nsstat <- tibble(wmis_cv = numeric(), pmis_cv = numeric())
for (i in 1:nrow(snp_nsgenes)) {
  #use dataframes, not lists
  snp_nspath <- snp_nspath %>%
    add_row(gene_name = snp_nsgenes$gene_name[i], pathway = fx_dndsNsPathway(snp_nsgenes$product[i]))
	snp_nsstat <- rbind(snp_nsstat, fx_dndsNsStats(snp_nsgenes$gene_name[i]))
}
snp_nsgenes <- left_join(snp_nsgenes, snp_nspath, by = c("gene_name"))
snp_nsgenes <- left_join(snp_nsgenes, snp_nsstat, by = c("gene_name"))
rm(snp_nspath, snp_nsstat)

@@

# Violin plot for NS genes with global mutations
snp_plot_nsgenes <- ggviolin(snp_nsgenes, x = "pathway", y = "wmis_cv", color = "pathway",
                             shape = "pathway", xlab = "Pathway", ylab = "dN/dS",
                             order = snp_sort_ns$pathway, add = "jitter", legend = "none",
                             palette = snp_sort_ns$color) +
  scale_y_continuous(limits = c(-0.2, 1.2)) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  stat_compare_means(method = "anova", label.y = 1.2) +
  stat_compare_means(comparisons = list(c("Copper", "Iron"), c("Sulfur", "Nickel")),
                     label = "p.format", method = "t.test")
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_nsgenes.pdf", snp_plot_nsgenes, height = 9, width = 7)

@@

# Lollipop plot of NS genes dN/dS x-axis range 0-2, y-axis PS genes vs selected CDS
snp_plot_nsgensub <- ggdotchart(snp_nsgenes_sub, x = "product", y = "dnds", color = "pathway",
                              xlab = "NS gene product", ylab = "dN/dS", y.text.col = FALSE,
                              rotate = TRUE, palette = snp_nsgenes_sub_col,
                              sorting = "descending", ggtheme = theme_pubr()) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2)
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_nsgenespath.pdf", snp_plot_nsgensub, height = 9, width = 10)

# Count total SNPs per mutation type for phylum
snp_topphyla <- snp_psgenes %>%
  dplyr::select(gene_name, n_syn, n_mis, n_non, bin, name) %>%
  dplyr::mutate(snps = n_syn + n_mis + n_non)
snp_topphyla <- snp_topphyla %>%
  dplyr::group_by(bin) %>%
  dplyr::summarise(sum_nsyn = sum(n_syn),
                   sum_nmis = sum(n_mis),
                   sum_nnon = sum(n_non))
snp_topphyla <- left_join(snp_topphyla, snp_dnds, by = c("bin")) %>%
  dplyr::select(-c("num", "genus", "species")) %>%
  pivot_longer(cols = c(sum_nsyn, sum_nmis, sum_nnon, predicted_genes),
               names_to = "muts", values_to = "nmuts")
#top 3 phyla and merge
snp_topphyla <- snp_topphyla %>%
  dplyr::filter(phylum == "Proteobacteria" | phylum == "Acidobacteriota" |
                  phylum == "Actinobacteriota") %>%
  dplyr::group_by(phylum, muts) %>%
  dplyr::summarise(tot = sum(nmuts), .groups = "drop")

# Barplot #snps/#genes
snp_plot_topphy <- ggplot(snp_topphyla, aes(x = reorder(phylum, -tot), y = tot, fill = muts)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  #geom_text(aes(label = tot), position = position_stack(reverse = TRUE), size = 2, vjust = 0.5) +
  scale_fill_manual(name = "", labels = c("gen", "mis", "non", "syn"),
                    values = c("#cba993", "#5db4b2", "#b687b6", "#2374a4")) +
  scale_y_continuous(limits = c(0, 70000)) +
  labs(x = "Phylum", y = "Number of PS mutations per type") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.9),
        strip.placement = "outside")
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_topphyla.pdf", snp_plot_topphy, height = 7, width = 5)

>>>>

suppressMessages(library(ape)) #tree basics
suppressMessages(library(treeio)) #join trees
suppressMessages(library(ggtree)) #tree plotting
suppressMessages(library(ggplot2)) #plots
suppressMessages(library(ggtreeExtra)) #geom_fruit
suppressMessages(library(ggnewscale)) #scales in ggplot2
suppressMessages(library(ggstar)) #geom_star for ggplot2
suppressMessages(library(tidyverse)) #strings, dataframes processing
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(tidytree)) #trees as dataframes
suppressMessages(library(RColorBrewer)) #color palettes
suppressWarnings(library(colorspace)) #colors manipulation
suppressWarnings(library(patchwork)) #side-by-side plotting
suppressWarnings(library(openxlsx)) #read excel
suppressWarnings(library(ComplexHeatmap)) #heatmap
suppressWarnings(library(PCAtools)) #pca
suppressWarnings(library(ggbiplot)) #pca plots
suppressWarnings(library(agricolae)) #fisher's test
suppressWarnings(library(ggpubr)) #stats & plots
suppressWarnings(library(dndscv)) #dN/dS of snps
suppressWarnings(library(clusterProfiler)) #gene enrichment
suppressWarnings(library(circlize)) #circular viz
```

@@

#sup 1
fig_s1 <- { plot_depth_top_phyla |
    (plot_depth_usgbs / plot_depth_spec + plot_layout(heights = c(0.85, 0.15))) } +
  plot_layout(ncol = 2, widths = c(0.25, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
fig_s1[[2]][[1]] <- fig_s1[[2]][[1]] + theme(axis.title.y = element_text(vjust = 0))
#fig_s1[[2]][[2]] <- fig_s1[[2]][[2]] + theme(axis.text.y = element_text(vjust = 0))
##use wrap_plots() to correct axis line/labels ??
ggsave("mags-all/fig_s1.pdf", fig_s1, height = 8, width = 16)
#ggsave("mags-all/fig_s1.png", fig_s1, height = 8, width = 16)
```

@@

# Sort ecosystems by median
anova_sort_cu <- kegg_bubble_cu %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(n = median(count_norm)) %>%
  dplyr::arrange(desc(n)) %>%
  as.data.frame()
anova_sort_fe <- kegg_bubble_fe %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(n = median(count_norm)) %>%
  dplyr::arrange(desc(n)) %>%
  as.data.frame()
anova_sort_s <- kegg_bubble_s %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(n = median(count_norm)) %>%
  dplyr::arrange(desc(n)) %>%
  as.data.frame()

@@

## Figure 2 & supplementary 2

```{r fig2}
# Merge plots using patchwork
fig_2 <- { { anova_boxplot_cu / anova_boxplot_fe +
    plot_layout(ncol = 1, widths = c(0.75)) } | anova_boxplot_s / kegg_bubbles_plot } +
  plot_annotation(tag_levels = list(c("a", "b", "c", "d"))) &
  theme(plot.tag =  element_text(face = "bold"))
#adjust y-label margin for sulfur plot
fig_2[[2]][[1]] <- fig_2[[2]][[1]] + theme(axis.title.y = element_text(vjust = -127))
ggsave("mags-all/fig_2.pdf", fig_2, height = 14, width = 18)
#ggsave("mags-all/fig_2.png", fig_2, height = 14, width = 18) #vjust=120
```

@@

```{r snps}
# Buildref for 1 bin (testing for bacteria species..)
#https://github.com/im3sanger/dndscv/issues/62
#buildref(cdsfile = "mags-all/5-SNPs/dndscv/buildref/cds_node_25854.txt",
#         genomefile = "mags-all/5-SNPs/dndscv/buildref/node_25854.fa",
#         outfile = "mags-all/5-SNPs/dndscv/buildref/testref.rda")
#load("mags-all/5-SNPs/dndscv/buildref/testref.rda")

# Buildref for all bins (very slow! ~2 hrs in building object)
#buildref(cdsfile = "mags-all/5-SNPs/dndscv/buildref/cds_table.txt",
#         genomefile = "mags-all/5-SNPs/dndscv/buildref/S15_bins.fa",
#         outfile = "mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda")

@@

# Explore output object..
#https://github.com/im3sanger/dndscv/issues/6
#https://github.com/im3sanger/dndscv/issues/64
head(snp_dndsout$sel_cv) #dN/dS ratios at gene-level
head(snp_dndsout$annotmuts) #list of annotated mutations
print(snp_dndsout$globaldnds) #driver globals

# Save matrix with dN/dS: driver discovery or genes under positive selection (PS)
snp_selcv <- snp_dndsout$sel_cv
rm(snp_dndsout)

# Group global dN/dS ratios:
#https://github.com/im3sanger/dndscv/issues/63
snp_global_ps <- snp_selcv %>%
  dplyr::filter(wmis_cv > 1) %>%
  dplyr::summarise(genes = n())
snp_global_ps$label = "dN/dS > 1"
snp_global_ns <- snp_selcv %>%
  dplyr::filter(wmis_cv < 1 & wmis_cv != 0) %>%
  dplyr::summarise(genes = n())
snp_global_ns$label = "dN/dS < 1"
snp_global_dnds <- rbind(snp_global_ps, snp_global_ns)
snp_global_dnds <- snp_global_dnds %>%
  dplyr::mutate(perc = round(genes * 100 / nrow(snp_selcv), 2)) %>% #add %
  dplyr::mutate(perc2 = paste0(genes, " (", perc, "%)")) #label to plot
rm(snp_global_ps, snp_global_ns) #clean memory

# Barplot with global dN/dS ratios (not used- for reference only*)
snp_plot_dndsglob <- ggplot(snp_global_dnds, aes(x = genes, y = reorder(label, genes), fill = label)) +
  geom_col(position = "dodge2", fill = "#4c6194") +
  geom_text(aes(label = perc2), vjust = 0.5, hjust = -0.1, size = 2) +
  scale_x_continuous(limits = c(0, 16000)) +
  labs(x = "Number of genes", y = "") +
  theme_classic() +
  theme(legend.position = "none")
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_global.pdf", snp_plot_dndsglob, height = 2, width = 5)

@@

# Get CDS annotations (pre-processed from Bakta results)
snp_refcds <- read.csv("mags-all/5-SNPs/dndscv/buildref/cds_table.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

@@

## FUNCTIONS ##
# Function that returns the contig where a gene was found
fx_dndsGene2Contig <- function(g) {
  c <- snp_refcds %>%
    dplyr::filter(gene.name == g) %>%
    dplyr::select(chr) %>%
    as.character()
  return(c)
}
# Function that returns the bin given a contig
fx_dndsContig2Bin <- function(c) {
  b <- snp_bins %>%
    dplyr::filter(contig == c) %>%
    dplyr::select(bin) %>%
    as.character()
  return(b)
}
# Function that gets gene names given an id
fx_dndsGeneName <- function(g) {
  g <- snp_genids %>%
    dplyr::filter(gene.id == g) %>%
    dplyr::select(gene.name) %>%
    as.character()
  return(g)
}
# Function that returns a row with the taxonomic classification of a bin
fx_dndsTaxaBin <- function(b) {
  t <- mag_samples %>%
    dplyr::mutate(bin = paste0(mag_name, ".", mag_id)) %>%
    dplyr::filter(bin == b) %>%
    dplyr::select(phylum, genus, species, completeness, contamination, predicted_genes)
  return(t)
}
# Function that process the product tag of NS genes
fx_dndsNsPathway <- function(p) {
  if (grepl("copper", p, ignore.case = TRUE) == TRUE)
    return("Copper")
  if (grepl("iron", p, ignore.case = TRUE) == TRUE)
    return("Iron")
  if (grepl("nickel", p, ignore.case = TRUE) == TRUE)
    return("Nickel")
  if (grepl("sulfur", p, ignore.case = TRUE) == TRUE)
    return("Sulfur")
  if (grepl("nitrogen", p, ignore.case = TRUE) == TRUE)
    return("Nitrogen")
}
# Function that gets dN/dS ratio and p-value for NS genes
fx_dndsNsStats <- function(g) {
  s <- snp_selcv %>%
    dplyr::filter(gene_name == g) %>%
    dplyr::select(gene_name, wmis_cv, pmis_cv)
  return(s)
}
## END FUNCTIONS ##

@@

snp_taxabins <- tibble(phylum = character(), genus = character(), species = character(),
                       completeness = numeric(), contamination = numeric(), predicted_genes = numeric())
for (i in 1:nrow(snp_dnds)) {
  #print(sprintf("%i -> bin: %s", i, fx_dndsTaxaBin(snp_dnds$bin[i])))
  snp_taxabins <- rbind(snp_taxabins, fx_dndsTaxaBin(snp_dnds$bin[i]))
}
snp_dnds['phylum'] <- snp_taxabins$phylum
snp_dnds['genus'] <- snp_taxabins$genus
snp_dnds['species'] <- snp_taxabins$species
snp_dnds['completeness'] <- snp_taxabins$completeness
snp_dnds['contamination'] <- snp_taxabins$contamination
snp_dnds['predicted_genes'] <- snp_taxabins$predicted_genes
rm(snp_taxabins)

@@

# Subset PS genes to plot mutations -> synonymous/missense (non-syn.)/nonsense
#discard HPs (by name)
#genes with at least 4 SNPs
#mark genes reaching significance (q < 0.05)
#mark genes: KJCAEPPB_04437 -> nickel pathway; NEBJFEFK_01910 -> sulfur pathway
snp_psgenes_sub <- snp_psgenes %>%
  dplyr::filter(nchar(name) < 14) %>%
  dplyr::mutate(snps = n_syn + n_mis + n_non) %>%
  dplyr::filter(snps >= 4) %>%
  dplyr::mutate(signif = ifelse(pmis_cv < 0.05, "*", "")) %>%
  dplyr::mutate(path = case_when(gene_name == "KJCAEPPB_04437" ~ "Ni",
                                 gene_name == "NEBJFEFK_01910" ~ "S", TRUE ~ "")) %>%
  distinct(name, .keep_all = TRUE) %>% #remove repeated names
  pivot_longer(cols = c(n_syn, n_mis, n_non), names_to = "mut", values_to = "n_mut")

# Save PS gene names and ids to file
if (file.exists("mags-all/5-SNPs/dndscv/results/ps_genes.txt") == FALSE) {
  write.table(snp_psgenes_sub %>% dplyr::select(gene_name) %>% distinct(gene_name), "mags-all/5-SNPs/dndscv/results/ps_genes.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}
if (file.exists("mags-all/5-SNPs/dndscv/results/kofam/ps_genes_ids.txt") == FALSE) {
  write.table(unique(snp_psgenes_sub$name), "mags-all/5-SNPs/dndscv/results/kofam/ps_genes_ids.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}

# Plot global driver genes by MAG/species
snp_plot_psbins <- ggplot(snp_dnds, aes(x = num, y = reorder(bin, num), fill = bin)) +
  geom_bar(stat = "identity", fill = "#E31A1C") +
  geom_text(aes(label = num), vjust = 0.5, hjust = -0.5, size = 2.5) +
  labs(x = "Number of genes under PS", y = "MAG Bin") +
  scale_x_continuous(limits = c(0, 255)) +
  theme_classic() +
  theme(strip.background = element_blank()) +
  facet_grid(.~species)
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psbins.pdf", snp_plot_psbins, height = 9, width = 10)

#add completeness & contamination barplots
snp_plot_psbins2 <- ggplot(snp_dnds, aes(x = completeness, y = reorder(bin, num), fill = bin)) +
  geom_col(fill = "#3c4d76") +
  geom_vline(xintercept = 100, linetype = 1, linewidth = 0.2) +
  labs(x = "Completeness", y = "") +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())
snp_plot_psbins3 <- ggplot(snp_dnds, aes(x = contamination, y = reorder(bin, num), fill = bin)) +
  geom_col(fill = "#763c4d") +
  geom_vline(xintercept = max(snp_dnds$contamination), linetype = 1, linewidth = 0.2) +
  labs(x = "Contamination", y = "") +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())

#merge
snp_plot_psbins <- wrap_plots(snp_plot_psbins, snp_plot_psbins2, snp_plot_psbins3,
                              nrow = 1, ncol = 3, widths = c(1, 0.15, 0.15))
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psbins.pdf", snp_plot_psbins, height = 9, width = 12)
rm(snp_plot_psbins2, snp_plot_psbins3)

@@

#merge
snp_plot_psgenes <- (snp_plot_psgenes | snp_plot_muts) +
  plot_layout(nrow = 1, ncol = 2, widths = c(0.55, 0.5))
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psgenmuts.pdf", snp_plot_psgenes, height = 11, width = 11)
rm(snp_plot_muts)

# Load negative selected (NS) genes from selected pathways
snp_nsgenes <- read.csv("mags-all/5-SNPs/dndscv/results/selected_cds_metals.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Infer pathway from product and add dndscv attributes
snp_nspath <- tibble(gene_name = character(), pathway = character())
snp_nsstat <- tibble(wmis_cv = numeric(), pmis_cv = numeric())
for (i in 1:nrow(snp_nsgenes)) {
  #use dataframes, not lists
  snp_nspath <- snp_nspath %>%
    add_row(gene_name = snp_nsgenes$gene_name[i], pathway = fx_dndsNsPathway(snp_nsgenes$product[i]))
	snp_nsstat <- rbind(snp_nsstat, fx_dndsNsStats(snp_nsgenes$gene_name[i]))
}
snp_nsgenes <- left_join(snp_nsgenes, snp_nspath, by = c("gene_name"))
snp_nsgenes <- left_join(snp_nsgenes, snp_nsstat, by = c("gene_name"))
rm(snp_nspath, snp_nsstat)

@@

# Violin plot for NS genes with global mutations
snp_plot_nsgenes <- ggviolin(snp_nsgenes, x = "pathway", y = "wmis_cv", color = "pathway",
                             shape = "pathway", xlab = "Pathway", ylab = "dN/dS",
                             order = snp_sort_ns$pathway, add = "jitter", legend = "none",
                             palette = snp_sort_ns$color) +
  scale_y_continuous(limits = c(-0.2, 1.3)) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  stat_compare_means(method = "anova", label.y = 1.3) +
  stat_compare_means(comparisons = list(c("Copper", "Iron"), c("Sulfur", "Nickel"),
                                        c("Copper", "Nickel")), label = "p.format", method = "t.test")
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_nsgenes.pdf", snp_plot_nsgenes, height = 9, width = 7)

@@

# Lollipop plot of NS genes dN/dS x-axis range 0-2, y-axis PS genes vs selected CDS
snp_plot_nsgensub <- ggdotchart(snp_nsgenes_sub, x = "product", y = "dnds", color = "pathway",
                              xlab = "NS gene product", ylab = "Mean dN/dS", y.text.col = FALSE,
                              rotate = TRUE, palette = snp_nsgenes_sub_col,
                              sorting = "descending", ggtheme = theme_pubr()) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2)
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_nsgenesprod.pdf", snp_plot_nsgensub, height = 9, width = 10)

# Global gene estimation of dn/ds (<1 & >1) per number of genes and phylum
snp_estim <- snp_selcv %>%
  dplyr::select(gene_name, wmis_cv, pmis_cv) %>%
  dplyr::filter(wmis_cv != 0)

# Populate global dataset (~2 min)
snp_gene2contig <- vector()
snp_contig2bin <- vector()
snp_genenames <- vector()
for (i in 1:nrow(snp_estim)) {
	snp_gene2contig <- append(snp_gene2contig, fx_dndsGene2Contig(snp_estim$gene_name[i]))
	snp_contig2bin <- append(snp_contig2bin, fx_dndsContig2Bin(snp_gene2contig[i]))
	snp_genenames <- append(snp_genenames, fx_dndsGeneName(snp_estim$gene_name[i]))
}
snp_estim['contig'] <- snp_gene2contig
snp_estim['bin'] <- snp_contig2bin
snp_estim['name'] <- snp_genenames
rm(snp_refcds, snp_bins, snp_genids, snp_gene2contig, snp_contig2bin, snp_genenames) #clean memory

# Add taxa & QC
snp_estim <- left_join(snp_estim, snp_dnds, by = c("bin")) %>%
  dplyr::select(-c("contig", "num", "genus", "species"))

# Normalize dN/dS per gene by MAG contamination, then group by phylum
snp_estim_phy <- snp_estim %>%
  dplyr::mutate(dnds_adj = wmis_cv * (1 - contamination / 100)) %>%
  dplyr::group_by(phylum, bin) %>%
  dplyr::summarise(snps = n(), dnds_sum = sum(dnds_adj)) %>%
  ungroup() %>%
  dplyr::mutate(dnds_glob = dnds_sum / snps) %>% #divide by total genes and top 3 phyla
  dplyr::filter(phylum == "Proteobacteria" | phylum == "Acidobacteriota" | phylum == "Actinobacteriota")

# Sort phylum by median and set color
snp_estim_sort_phy <- snp_estim_phy %>%
  dplyr::group_by(phylum) %>%
  dplyr::summarise(n = median(dnds_glob)) %>%
  dplyr::arrange(desc(n))
snp_estim_sort_phy$color <- rep("#4c6194", 3)
